\vspace{-\baselineskip} Fast and scalable Gaussian process modeling with applications to astronomical time series \vspace{-2\baselineskip}

### 2017-05-31 paper/ms.tex
Added an example of the semi-separable matrix decomposition

The relationship between the \celeriteterm\ model and semi-separable linear
algebra was first recognized by Ambikasaran \citep{Ambikasaran:2015}.
A rank-$R$ semi-separable matrix is a matrix $K$ where the elements can be
written as
\begin{eqnarray}
K_{n,m} = \left\{\begin{array}{ll}
    \sum_{r=1}^R U_{n,r}\,V_{m,r} & \mathrm{if}\,m<n \\
    A_{n,n} & \mathrm{if}\,m=n \\
    \sum_{r=1}^R U_{m,r}\,V_{n,r} & \mathrm{otherwise}
\end{array}\right.
\end{eqnarray}
for some $N \times R$-dimensional matrices $U$ and $V$, and an $N \times
N$-dimensional diagonal matrix $A$.
This definition can be equivalently written as
\begin{eqnarray}\eqlabel{semi-sep}
K = A + \mathrm{tril}(U\,V^\T) + \mathrm{triu}(V\,U^\T)
\end{eqnarray}
where the $\mathrm{tril}$ function extracts the strict lower triangular
component of its argument and $\mathrm{triu}$ is the equivalent strict upper
triangular operation.
For the \celeriteterm\ model described in the previous section, the rank of
the semi-separable matrix is $R = 2\,J$ and the components of the relevant
matrices are
\begin{eqnarray}
U_{n,2\,j-1} &=& a_j\,e^{-c_j\,t_n}\,\cos(d_j\,t_n) +
    b_j\,e^{-c_j\,t_n}\,\sin(d_j\,t_n) \\
U_{n,2\,j} &=& a_j\,e^{-c_j\,t_n}\,\sin(d_j\,t_n) -
    b_j\,e^{-c_j\,t_n}\,\cos(d_j\,t_n) \\
V_{m,2\,j-1} &=& e^{c_j\,t_m}\,\cos(d_j\,t_m) \\
V_{m,2\,j} &=& e^{c_j\,t_m}\,\sin(d_j\,t_m) \\
A_{n,n} &=& \sigma_n^2 + \sum_{j=1}^J a_j \quad.
\end{eqnarray}

>>>>

The relationship between the \celeriteterm\ model and semi-separable linear
algebra was first recognized by Ambikasaran \citep{Ambikasaran:2015}.
A rank-$R$ semi-separable matrix is a matrix $K$ where the elements can be
written as
\begin{eqnarray}
K_{n,m} = \left\{\begin{array}{ll}
    \sum_{r=1}^R U_{n,r}\,V_{m,r} & \mathrm{if}\,m<n \\
    A_{n,n} & \mathrm{if}\,m=n \\
    \sum_{r=1}^R U_{m,r}\,V_{n,r} & \mathrm{otherwise}
\end{array}\right.
\end{eqnarray}
for some $N \times R$-dimensional matrices $U$ and $V$, and an $N \times
N$-dimensional diagonal matrix $A$.
This definition can be equivalently written as
\begin{eqnarray}\eqlabel{semi-sep}
K = A + \mathrm{tril}(U\,V^\T) + \mathrm{triu}(V\,U^\T)
\end{eqnarray}
where the $\mathrm{tril}$ function extracts the strict lower triangular
component of its argument and $\mathrm{triu}$ is the equivalent strict upper
triangular operation.  Note that the last two terms in this equation are
transposes of one another.
For the \celeriteterm\ model described in the previous section, the rank of
the semi-separable matrix is $R = 2\,J$ and the components of the relevant
matrices are
\begin{eqnarray}
U_{n,2\,j-1} &=& a_j\,e^{-c_j\,t_n}\,\cos(d_j\,t_n) +
    b_j\,e^{-c_j\,t_n}\,\sin(d_j\,t_n) \\
U_{n,2\,j} &=& a_j\,e^{-c_j\,t_n}\,\sin(d_j\,t_n) -
    b_j\,e^{-c_j\,t_n}\,\cos(d_j\,t_n) \\
V_{m,2\,j-1} &=& e^{c_j\,t_m}\,\cos(d_j\,t_m) \\
V_{m,2\,j} &=& e^{c_j\,t_m}\,\sin(d_j\,t_m) \\
A_{n,n} &=& \sigma_n^2 + \sum_{j=1}^J a_j \quad.
\end{eqnarray}

To help clarify these equations, we write out here an example of a semi-separable matrix with a 
single \celeriteterm\ term with $N=3$ and $c=d=1$:
\begin{eqnarray}\eqlabel{semi-separable-structure}
K &=& \\
&&\begin{pmatrix}
%    {\scriptscriptstyle a+{\sigma_1}^2 }& {\scriptscriptstyle e^{-c\,\tau_{2,1}}\left(a\,\cos{(d \tau_{2,1})}+b\,\sin{(d \tau_{2,1})}\right)} & {\scriptscriptstyle e^{-c\,\tau_{3,1}}\left(a\,\cos{(d \tau_{3,1})}+b\,\sin{(d \tau_{3,1})}\right)}\\ 
%    {\scriptscriptstyle e^{-c\,\tau_{2,1}}\left(a\,\cos{(d \tau_{2,1})}+b\,\sin{(d \tau_{2,1})}\right)}&{\scriptscriptstyle a+{\sigma_2}^2 }& {\scriptscriptstyle e^{-c\,\tau_{3,2}}\left(a\,\cos{(d \tau_{3,2})}+b\,\sin{(d \tau_{3,2})}\right)}\\
%    {\scriptscriptstyle e^{-c\,\tau_{3,2}}\left(a\,\cos{(d \tau_{3,2})}+b\,\sin{(d \tau_{3,2})}\right)}&{\scriptscriptstyle  e^{-c\,\tau_{3,2}}\left(a\,\cos{(d \tau_{3,2})}+b\,\sin{(d \tau_{3,2})}\right)} & \scriptscriptstyle{a + {\sigma_3}^2}
    {\scriptscriptstyle a+{\sigma_1}^2 }& {\scriptscriptstyle e^{-\tau_{2,1}}\left(a\,\cos{\tau_{2,1}}+b\,\sin{\tau_{2,1}}\right)} & {\scriptscriptstyle e^{-\tau_{3,1}}\left(a\,\cos{\tau_{3,1}}+b\,\sin{\tau_{3,1}}\right)}\\ 
    {\scriptscriptstyle e^{-\tau_{2,1}}\left(a\,\cos{\tau_{2,1}}+b\,\sin{\tau_{2,1}}\right)}&{\scriptscriptstyle a+{\sigma_2}^2 }& {\scriptscriptstyle e^{-\tau_{3,2}}\left(a\,\cos{\tau_{3,2}}+b\,\sin{\tau_{3,2}}\right)}\\
    {\scriptscriptstyle e^{-\tau_{3,2}}\left(a\,\cos{\tau_{3,2}}+b\,\sin{\tau_{3,2}}\right)}&{\scriptscriptstyle  e^{-\tau_{3,2}}\left(a\,\cos{\tau_{3,2}}+b\,\sin{\tau_{3,2}}\right)} & \scriptscriptstyle{a + {\sigma_3}^2}
\end{pmatrix}\\
&=&
\begin{pmatrix}
    {\scriptscriptstyle a+{\sigma_1}^2} & {\scriptscriptstyle 0} & {\scriptscriptstyle 0} \\
    {\scriptscriptstyle 0} & {\scriptscriptstyle a+{\sigma_2}^2 }& {\scriptscriptstyle 0} \\
    {\scriptscriptstyle 0} & {\scriptscriptstyle 0} & {\scriptscriptstyle a + {\sigma_3}^2}
\end{pmatrix}\\
&+&\mathrm{tril}\left[\begin{pmatrix}
%    {\scriptscriptstyle e^{-c\,t_1}\left(a\,\cos{(d t_1)}+b\,\sin{(d t_1)}\right)}&
%    {\scriptscriptstyle e^{-c\,t_1}\left(a\,\sin{(d t_1)}-b\,\cos{(d t_1)}\right)}\\
%    {\scriptscriptstyle e^{-c\,t_2}\left(a\,\cos{(d t_2)}+b\,\sin{(d t_2)}\right)}&
%    {\scriptscriptstyle e^{-c\,t_2}\left(a\,\sin{(d t_2)}-b\,\cos{(d t_2)}\right)}\\
%    {\scriptscriptstyle e^{-c\,t_3}\left(a\,\cos{(d t_3)}+b\,\sin{(d t_3)}\right)}&
%    {\scriptscriptstyle e^{-c\,t_3}\left(a\,\sin{(d t_3)}-b\,\cos{(d t_3)}\right)}
    {\scriptscriptstyle e^{-t_1}\left(a\,\cos{t_1}+b\,\sin{t_1}\right)}&
    {\scriptscriptstyle e^{-t_1}\left(a\,\sin{t_1}-b\,\cos{t_1}\right)}\\
    {\scriptscriptstyle e^{-t_2}\left(a\,\cos{t_2}+b\,\sin{t_2}\right)}&
    {\scriptscriptstyle e^{-t_2}\left(a\,\sin{t_2}-b\,\cos{t_2}\right)}\\
    {\scriptscriptstyle e^{-t_3}\left(a\,\cos{t_3}+b\,\sin{t_3}\right)}&
    {\scriptscriptstyle e^{-t_3}\left(a\,\sin{t_3}-b\,\cos{t_3}\right)}
\end{pmatrix} %\right.\\
%&\times &
\times
%\left.
\begin{pmatrix}
% {\scriptscriptstyle e^{c\,t_1} \cos{(d t_1)}} &{\scriptscriptstyle e^{c\,t_2} \cos{(d t_2)}} & {\scriptscriptstyle e^{c\,t_3} \cos{(d t_3)}}\\
% {\scriptscriptstyle e^{c\,t_1} \sin{(d t_1)}} & {\scriptscriptstyle e^{c\,t_2} \sin{(d t_2)}} & {\scriptscriptstyle e^{c\,t_3} \sin{(d t_3)}}
% {\scriptscriptstyle e^{t_1} \cos{t_1}} & {\scriptscriptstyle e^{t_2} \cos{t_2}} & {\scriptscriptstyle e^{t_3} \cos{t_3}}\\
% {\scriptscriptstyle e^{t_1} \sin{t_1}} & {\scriptscriptstyle e^{t_2} \sin{t_2}} & {\scriptscriptstyle e^{t_3} \sin{t_3}}
 {\scriptscriptstyle e^{t_1} \cos{t_1}} & {\scriptscriptstyle e^{t_1} \sin{t_1}} \\
 {\scriptscriptstyle e^{t_2} \cos{t_2}} & {\scriptscriptstyle e^{t_2} \sin{t_2}}\\
 {\scriptscriptstyle e^{t_3} \cos{t_3}} & {\scriptscriptstyle e^{t_3} \sin{t_3}}
\end{pmatrix}^T\right]\\
&+&
\mathrm{triu}\left[\begin{pmatrix}
% {\scriptscriptstyle e^{c\,t_1} \cos{(d t_1)}} & {\scriptscriptstyle e^{c\,t_1} \sin{(d t_1)}} \\
% {\scriptscriptstyle e^{c\,t_2} \cos{(d t_2)}} & {\scriptscriptstyle e^{c\,t_2} \sin{(d t_2)}}\\
% {\scriptscriptstyle e^{c\,t_3} \cos{(d t_3)}} & {\scriptscriptstyle e^{c\,t_3} \sin{(d t_3)}}
 {\scriptscriptstyle e^{t_1} \cos{t_1}} & {\scriptscriptstyle e^{t_1} \sin{t_1}} \\
 {\scriptscriptstyle e^{t_2} \cos{t_2}} & {\scriptscriptstyle e^{t_2} \sin{t_2}}\\
 {\scriptscriptstyle e^{t_3} \cos{t_3}} & {\scriptscriptstyle e^{t_3} \sin{t_3}}
\end{pmatrix} % \right.\\
\times
%\left.
\begin{pmatrix}
%    {\scriptscriptstyle e^{-c\,t_1}\left(a\,\cos{(d t_1)}+b\,\sin{(d t_1)}\right)}&
%    {\scriptscriptstyle e^{-c\,t_2}\left(a\,\cos{(d t_2)}+b\,\sin{(d t_2)}\right)}&
%    {\scriptscriptstyle e^{-c\,t_3}\left(a\,\cos{(d t_3)}+b\,\sin{(d t_3)}\right)}\\
%    {\scriptscriptstyle e^{-c\,t_1}\left(a\,\sin{(d t_1)}-b\,\cos{(d t_1)}\right)}&
%    {\scriptscriptstyle e^{-c\,t_2}\left(a\,\sin{(d t_2)}-b\,\cos{(d t_2)}\right)}&
%    {\scriptscriptstyle e^{-c\,t_3}\left(a\,\sin{(d t_3)}-b\,\cos{(d t_3)}\right)}
%    {\scriptscriptstyle e^{-t_1}\left(a\,\cos{t_1}+b\,\sin{t_1}\right)}&
%    {\scriptscriptstyle e^{-t_2}\left(a\,\cos{t_2}+b\,\sin{t_2}\right)}&
%    {\scriptscriptstyle e^{-t_3}\left(a\,\cos{t_3}+b\,\sin{t_3}\right)}\\
%    {\scriptscriptstyle e^{-t_1}\left(a\,\sin{t_1}-b\,\cos{t_1}\right)}&
%    {\scriptscriptstyle e^{-t_2}\left(a\,\sin{t_2}-b\,\cos{t_2}\right)}&
%    {\scriptscriptstyle e^{-t_3}\left(a\,\sin{t_3}-b\,\cos{t_3}\right)}
    {\scriptscriptstyle e^{-t_1}\left(a\,\cos{t_1}+b\,\sin{t_1}\right)}&
    {\scriptscriptstyle e^{-t_1}\left(a\,\sin{t_1}-b\,\cos{t_1}\right)}\\
    {\scriptscriptstyle e^{-t_2}\left(a\,\cos{t_2}+b\,\sin{t_2}\right)}&
    {\scriptscriptstyle e^{-t_2}\left(a\,\sin{t_2}-b\,\cos{t_2}\right)}\\
    {\scriptscriptstyle e^{-t_3}\left(a\,\cos{t_3}+b\,\sin{t_3}\right)}&
    {\scriptscriptstyle e^{-t_3}\left(a\,\sin{t_3}-b\,\cos{t_3}\right)}
\end{pmatrix}^T\right]\\
&=& A + \mathrm{tril}(U V^T) + \mathrm{triu}(V U^T).
\end{eqnarray}
Note that rather than storing the full $N \times N$ covariance matrix, it
is only necessary to store the $2J \times N$ matrices $U$ and $V$, as well
as the diagonal terms, for a total storage volume of $(4J+1) \times N$ floating
point numbers.
